---
title: "Alcohol Consumption in Schools"
author: "Hypothesis Heroes - Drew Davison, Lisa Zhang, Ellie Culman, Austin Chang"
date: "11/14/23"
format: pdf
execute: 
  warning: false
  message: false
  echo: false
editor: visual
---

```{r}
#| label: load packages and data
#| message: FALSE
library(tidyverse)
library(tidymodels)
library(kableExtra)
library(patchwork)
library(MASS)
library(knitr)
library(rms)
```

## Introduction and data

All around the world, underage drinking among students is a significant public health issue and takes a huge toll on the quality of students' lives and education. There is plenty of existing documentation on the effects of socioeconomic status on risky drinking behavior amongst college students. For example, a paper by Susan E. Collins (Collins SE, 2016), PhD, professor and licensed clinical psychologist, found that individuals with lower socioeconomic status as well as people of racial and ethnic minorities and homelessness experience greater alcohol-related consequences. Additionally, studies from the National Institutes of Health (NIAAA, 2023) have found that "aspects of college life---such as unstructured time, widespread availability of alcohol, inconsistent enforcement of underage drinking laws, and limited interactions with parents and other adults" drives up rates of underage drinking, and "college students have higher binge-drinking rates and a higher incidence of driving under the influence of alcohol than their noncollege peers." Since most existing literature on underage student drinking focuses on college students, we wanted to examine the factors contributing to drinking amongst secondary school students. Our research question is as follows:

How do social indicators affect student alcohol consumption in secondary schools?

We hypothesize that factors like gender, familial status, family and school support, and other social and economic indicators will strongly influence the rates that secondary school students consume alcohol and that increased alcohol consumption is correlated with their school performance.

```{r}
alc <- read_csv("./data/student-por.csv")
```

This data is from a Kaggle public dataset, originally sourced from the UC Irvine Machine Learning Repository. The data consists of information collected in 2008 on secondary school students from two schools in Portugal: Gabriel Pereira and Mousinho da Silveira. There are 649 observations, each one being a student, and 33 variables which cover a range of characteristics about each student's family, education, social situation, alcohol consumption, and grades in their Portuguese language class. The data were collected through a survey of each student in the class. Some key variables we will be examining are sex, studytime, absences, Medu, Fedu, famrel, freetime, goout, health, Fjob, Mjob, famsize, reason, nursery. Our response variable is Walc, which is the students' average weekend alcohol consumption, on a scale of 1-5, with 5 being very high.

```{r}
#| fig-height: 10
#| fig-width: 15

walchist <- alc |>
  ggplot(aes(x = Walc)) +
  geom_histogram(stat = "count") +
  labs(x = "Weekend Alcohol Consumption",
       y = "Frequency",
       title = "Distribution of Weekend Alcohol Consumption Scores",
       subtitle = "On a scale of 1 (very low) to 5 (very high)")

alcline <- alc %>%
  ggplot(aes(x = Dalc)) +
  geom_line(aes(y = ..count.., color = "Dalc"), stat = 'count') +
  geom_line(aes(x = Walc, y = ..count.., color = "Walc"), stat = 'count') +
  labs(x = "Alcohol Consumption",
       y = "Frequency",
       color = "Weekday or Weekend",
       title = "Distribution of Alcohol Consumption Scores Weekday vs. Weekend",
       subtitle = "On a scale of 1 (very low) to 5 (very high)") +
  scale_color_manual(
    values = c("Dalc" = "blue", "Walc" = "red"),
    labels = c("Weekday", "Weekend")
  )
 
walcstudy <- alc |>
  ggplot(aes(x = studytime, y = Walc)) +
  geom_point() +
  geom_jitter() +
  geom_smooth(method = "lm") +
  labs(x = "Study Time",
       y = "Weekend Alcohol Consumption Scores",
       title = "Study Time vs. Weekend Alcohol Consumption Scores",
       subtitle = "On a scale of 1 to 4: 1 = <2 hours, 2 = 2-5 hours, 3 = 5-10 hours, 4 = >10 hours")

walcfamsup <- alc |>
  ggplot(aes(x = famsup, y = Walc)) +
  geom_boxplot() +
  labs(x = "Family Educational Support (Yes or No)",
       y = "Weekend Alcohol Consumption Scores",
       title = "Family Educational Support vs. Weekend Alcohol Consumption Scores")

(walchist | alcline) / (walcstudy | walcfamsup)
```

### Distribution of Response Variable

```{r}
summary(alc$Walc) |>
  tidy() |>
  kable()

alc_percentage <- alc %>%
    group_by(Walc) %>%
    summarise(count = n()) %>%
    mutate(Percentage = sprintf("%.2f%%", (count / sum(count) * 100)))

alc_percentage |>
  kable()
```

Our initial EDA looked at the distributions of our response variable, weekend alcohol consumption, along with some other relevant factors like study time, family educational support, and weekday alcohol consumption that we hypothesized would have some correlation with weekend alcohol consumption. We found that the majority of students reported their weekend drinking as very low (1) and the amounts reporting subsequent numbers of 2, 3, 4, 5 gradually decreased in number. We also saw that more students reported higher levels of drinking on the weekends (2-5). We also observed a negative linear relationship between weekend alcohol consumption and study time as well as higher weekend alcohol consumption for students with no family educational support vs. those with family educational support.

## Methodology

We are using a multivariable linear regression to examine the effects of the predictor variables sex, Pstatus, schoolsup, famsup, Mjob, Fjob, freetime, studytime, absences, and failures on the response variable Walc, or weekend alcohol consumption. We are using a linear model rather than logistic since Walc is not binary. With many behavioral or otherwise psychological models, the main focus is to identify the greatest effects rather than any effects. As such, because we weren't sure which predictors would give us significant effects, we started by using the stepAIC function to perform a backwards selection on the full model which allowed us to narrow down the predictors with significant effects. The backwards selection process takes the full model and then calculated the AIC of the full model without each of the variables. From this list of models (the full model and the several full models missing one variable), it removes the variable associated with the lowest AIC. This process is then repeated with the new full model, now with one less predictor, and continues repeating until the full model has the lowest AIC. This ensures that all the remaining predictors are statistically significant at a level of p < 0.05. For instance, in our backwards selection, the model that removed "guardian" as a predictor had the lowest AIC out of all the models so selection continued with a new full model consisting of all predictors other than guardian. We then ran a VIF function to determine multicollinearity and found that our predictors were satisfactorily independent of one another. We considered interactions between studytime and absences, freetime and goout, as well as famrel and famsize. These interaction were chosen based on their real-life contexts (for instance, the effect of family size would likely change depending on familial relationship). The interactions did not significantly improve the r-squared or rmse values as well as greatly increasing collinearity. As such, we selected Model 1 as our model.

In our model we also considered including interactions between freetime and studytime, failures and studytime, failures and absences, as well as famsup and Mjob and famsup and Fjob. In our model we used step_dummy() to create dummy variables for all categorical variables, step_interact() to create all of our interaction variables, and step_zv() to remove all variables with only one value.

The StepAIC and VIF steps can be found in the appendix.

### Split Data into Testing and Training Data

```{r}
#| label: split-data
set.seed(210)
alc_split <- initial_split(alc, prop = 0.75)
alc_train <- training(alc_split)
alc_test <- testing(alc_split)
```

### Split Training Data in 5 Folds for Cross-Validation

```{r}
set.seed(210)
folds <- vfold_cv(alc_train, v = 5) 

```

### Model 1

```{r}
alc_rec1 <- recipe(Walc ~ sex + Fjob + studytime + absences + Medu + 
                     Fedu + Mjob + famsize + reason + nursery +
                     famrel + freetime + goout + health, data = alc_train) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_center(absences)

alc_rec1
```

#### Model Specification + Workflow 1

```{r}
alc_spec1 <- linear_reg() |>
  set_engine("lm")

alc_wflow1 <- workflow() |>
  add_model(alc_spec1) |>
  add_recipe(alc_rec1) 

alc_fit1 <- alc_wflow1 |>
  fit(data = alc_train)

tidy(alc_fit1) |>
  kable(digits = 3)


```

#### RMSE and R-Squared from 5-Fold Cross Validation 1

```{r}
calc_model_stats <- function(x) {
  glance(extract_fit_parsnip(x)) |>
    select(adj.r.squared, AIC, BIC)
}

alc_fit_cv1 <- alc_wflow1 |>
  fit_resamples(resamples = folds,
                control = control_resamples(extract = calc_model_stats))

collect_metrics(alc_fit_cv1, summarize = TRUE) |>
  kable(digits = 3)
```

### Model 2 with Interactions

```{r}
#| label: create-recipe
alc_rec2 <- recipe(Walc ~ sex + Fjob + studytime + absences + Medu + 
                     Fedu + Mjob + famsize + reason + nursery +
                     famrel + freetime + goout + health, data = alc_train) |>
  step_dummy(all_nominal_predictors()) |>
  step_interact(terms = ~ starts_with("studytime"):starts_with("absences")) |>
  step_interact(terms = ~ starts_with("freetime"):starts_with("goout")) |>
  step_interact(terms = ~ starts_with("famsize"):starts_with("famrel")) |>
  step_center(absences) |>
  step_zv(all_predictors())

alc_rec2
```

#### Model Specification + Workflow 2

```{r}
alc_spec <- linear_reg() |>
  set_engine("lm")

alc_wflow2 <- workflow() |>
  add_model(alc_spec) |>
  add_recipe(alc_rec2) 

alc_fit2 <- alc_wflow2 |>
  fit(data = alc_train)

tidy(alc_fit2) |>
  kable(digits = 3)


```

#### RMSE and R-Squared from 5-Fold Cross Validation 2

```{r}
calc_model_stats <- function(x) {
  glance(extract_fit_parsnip(x)) |>
    select(adj.r.squared, AIC, BIC)
}

alc_fit_cv2 <- alc_wflow2 |>
  fit_resamples(resamples = folds,
                control = control_resamples(extract = calc_model_stats))

collect_metrics(alc_fit_cv2, summarize = TRUE) |>
  kable(digits = 3)
```

### Applying Model 1 to Testing Data

```{r}
alc_test_pred <- predict(alc_fit1, alc_test) |>
  bind_cols(alc_test)

rmse(alc_test_pred, truth = Walc, estimate = .pred) |>
  kable(digits = 3)
rsq(alc_test_pred, truth = Walc, estimate = .pred) |>
  kable(digits = 3)
```

### Checking Conditions for Selected Model

```{r}
alcfit_resid <- predict(alc_fit1, alc_train) |> 
  bind_cols(alc_train) |> 
  mutate(resid = Walc - .pred) 

alcfit_resid |> 
  ggplot(aes(x = .pred, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") + 
  labs(x = "Predicted Values", 
       y = "Residuals", 
       title = "Residuals vs. Predicted Values") 
```

```{r}
#| fig-height: 15 
#| fig-width: 15 
sex <- alcfit_resid |> 
  ggplot(aes(x = sex, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") + 
  labs(x = "Student's Sex", y = "Residuals", title = "Residuals vs. Student's Sex") 

fjob <- alcfit_resid |> 
  ggplot(aes(x = Fjob, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") + 
  labs(x = "Father Job", y = "Residuals", title = "Residuals vs. Father Job") 

mjob <- alcfit_resid |> 
  ggplot(aes(x = Mjob, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") + 
  labs(x = "Mother Job", y = "Residuals", title = "Residuals vs. Mother Job") 

studytime <- alcfit_resid |> 
  ggplot(aes(x = studytime, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") + 
  labs(x = "Amount of Studytime", y = "Residuals", title = "Residuals vs. Amount of Studytime") 

absences <- alcfit_resid |> 
  ggplot(aes(x = absences, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") + 
  labs(x = "Number of Absences", y = "Residuals", title = "Residuals vs. Number of Absences") 

medu <- alcfit_resid |> 
  ggplot(aes(x = Medu, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") + 
  labs(x = "Mother Education", y = "Residuals", title = "Residuals vs. Mother Education") 

fedu <- alcfit_resid |> 
  ggplot(aes(x = Fedu, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") + 
  labs(x = "Father Education", y = "Residuals", title = "Residuals vs. Father Education") 

famsize <- alcfit_resid |> 
  ggplot(aes(x = famsize, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") + 
  labs(x = "Family Size (Greater than 3 or Less than 3)", y = "Residuals", title = "Residuals vs. Family Size") 

reason <- alcfit_resid |> 
  ggplot(aes(x = reason, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") + 
  labs(x = "Reason for Drinking", y = "Residuals", title = "Residuals vs. Reason") 

nursery <- alcfit_resid |> 
  ggplot(aes(x = nursery, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") + 
  labs(x = "If Student went to Nursery School", y = "Residuals", title = "Residuals vs. Nursery") 

famrel <- alcfit_resid |> 
  ggplot(aes(x = famrel, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") + 
  labs(x = "Family Relationship Rating", y = "Residuals", title = "Residuals vs. Family Relationship") 

freetime <- alcfit_resid |> 
  ggplot(aes(x = freetime, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") + 
  labs(x = "Amount of Freetime", y = "Residuals", title = "Residuals vs. Freetime") 

goout <- alcfit_resid |> 
  ggplot(aes(x = goout, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") + 
  labs(x = "Amoung of Going Out", y = "Residuals", title = "Residuals vs. Go Out") 

health <- alcfit_resid |> 
  ggplot(aes(x = health, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") + 
  labs(x = "Student's Health", y = "Residuals", title = "Residuals vs. Health") 

(sex | fjob | mjob ) / (studytime | absences | famsize) / (medu | fedu | reason) / (nursery | famrel | freetime) / (goout | health)
```

```{r}
alcfit_resid |> 
  ggplot(aes(x = resid)) +
  geom_histogram(binwidth = 0.5) + 
  labs(x = "Predicted Values", y = "Residuals", title = "Distribution of Residuals")
```

Linearity: Linearity is met because there are no discernible patterns in the residuals plot of the predicted values and all of the individual predictors; the points are scattered relatively randomly.

Constant Variance: Constant Variance is met because in the bulk of the data, the vertical spread is relatively constant throughout; there does not appear to be a fan shape as you move from left to right.

Normality: Normality is satisfied because the distribution of residuals is approximately unimodal and symmetric.

Independence: Independence is satisfied because the data is a random sample, and we can reasonably assume that the data for one student does not effect the data for the other students.

## Results

Our model assumes independence between predictors, linearity, and the other assumptions of a multilinear regression model. These assumptions were supported by our diagnostic tests as well as the VIF test. Our final R squared was .224 and our final RMSE was 1.121. This suggests that our model accounts for around 22% of the total variation in weekend alcohol consumption with an error of about 1.12. While the R-squared is relatively low, we believe this is still a good model given the context we are trying to predict. Predicting teen behavior is a difficult task, as there are many random factors that can attribute to decision making. Our model only covers a select few, and even with other predicting factors, teen brains usually do not align with predicable patterns.

From the significant p-values, we concluded that the following variables had an effect on weekend alcohol consumption. As each level of Mother's education increases by one unit the expected weekend alcohol consumption score increases by 0.345 on average. As each level of quality of family relationship increases by one unit the expected weekend alcohol consumption score decreases by 0.307 on average. As each level of going out with friends increases by one unit the expected weekend alcohol consumption score increases by 0.381 on average. If the individual is Male, the expected weekend alcohol consumption score increases by 0.838 compared to the baseline of Female on average. If the individual's Mother works in the civil services industry, the expected weekend alcohol consumption score decreases by 0.737 compared to the baseline of stay at home Mom on average.

### Conclusion

From our data analysis, we conclude that family relationships has the biggest impact on weekend alcohol consumption. Extrapolating from this, we emphasize the importance of strong family dynamics for school aged children. This drives us to consider more research into what makes children feel supported by their families and how parents' employment impacts daily decisions. Furthermore, the mother's education level also has a significant impact on weekend alcohol consumption, and a potential future research question could examine this relationship. Our initial hypothesis is that as the mother's education level increases she is more likely to be working and out of the home compared to a mother with a lower level of education. Pending further research, we would continue to emphasize the role family relationships have on weekend alcohol consumption, especially for parents who spend less time in the home. The going out variable also revealed a significant positive relationship with weekend alcohol consumption which reveals that the secondary students who spend time going out with friends are more likely to drink alcohol on the weekends, potentially with their friends. We also found it interesting that men are more likely to participate in alcohol consumption at school age rather than women. We believe that may have something to local gender norms and allows for further research into young male substance abuse and mental health.

Our model is significantly limited by the data collection. The data comes from two schools in Portugal, making it unrealistic to apply to American teens, especially given the different alcohol laws in United States. Portugal also has different family dynamics and values than the US. Our data collection also is self reported on a scale of 1 to 5. Using a 1-5 scale limits our numerical analysis because every evaluation is relative to each participant's understanding of their own situation. This is why we also saw a trend towards the middle, since most participants are likely to pick a less extreme answer in a self reported context. If we were to replicate this analysis, we would seek to find data from the United States that has similar predictor variables, but ones that a measured quantitatively instead of on a 1-5 scale. An example of this would be estimating how many drinks consumed on the weekend rather than self reporting on the scale. We are satisfied with the categorical variables.

## Citations

Collins SE. Associations Between Socioeconomic Factors and Alcohol Outcomes. Alcohol Res. 2016;38(1):83-94. PMID: 27159815; PMCID: PMC4872618.

"Underage Drinking in the United States (Ages 12 to 20)." *National Institute on Alcohol Abuse and Alcoholism*, U.S. Department of Health and Human Services,niaaa.nih.gov

## Appendix

### StepAIC Model

```{r}
#|label: stepAIC-model
lm1 <- lm(Walc ~ .-Dalc, data = alc_train)
stepAIC(lm1, direction = "backward",trace = 0) |>
  tidy() |>
  kable(digits = 3)
```

#### VIF Multicollinearity Test 1

```{r}
alc_fit_model_1 <- extract_fit_parsnip(alc_fit1)
vif(alc_fit_model_1$fit) |>
  tidy() |>
  kable(digits = 3)
```

#### VIF Multicollinearity Test 2

```{r}
alc_fit_model_2 <- extract_fit_parsnip(alc_fit2)
vif(alc_fit_model_2$fit) |>
  tidy() |>
  kable(digits = 3)
```
#### Fit To Testing Data
```{r}
alc_fit <- alc_wflow1 |>
  fit(data = alc_test)

tidy(alc_fit)|>
  kable(digits = 3)
```